---
title: "02_model"
author: "Riley Butler"
date: "2026-08-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This file is used for model construction. 

```{r}
# set-up 
if (!dir.exists("results")) {
  dir.create("results", recursive = TRUE)
}

wgcna_path <- "data/wgcna_results.rds"
if (!file.exists(processed_path)) {
  system("Rscript 01_wgcna.R")
}
```

```{r}
# load libraries 
library(tidyverse)
library(performance)
```

```{r}
# load data
data <- readRDS(wgcna_path)
clinical_factors <- data$pheno
eigengenes <- data$eigengenes

# create dataframe
model_data <- data.frame(clinical_factors,eigengenes)

# swapping variables - maybe do this early on? 
model_data$pathologic_response <- ifelse(model_data$pathologic_response == 0, 1, 0)

```

Now that the data is set-up - lets explore! 

# Logistic Regression Models 

These models have assumptions, including 
1. The outcome is binary. In this case, pCR = 1 and RD = 0
2. Linearity of log-odds. This means continuous predictors (MEs, age, etc) are linearly related to the log-odds of the outcome. 
3. No multicolinearity, the predictors aren't strongly correlated. This is measured by VIF, with a VIF less than 10 indicating no colinearity.  
```{r}

# null model 
m0 <- glm(pathologic_response ~ 1, data = model_data, family = binomial)
summary(m0)

# full model - eigengenes
m1 <- glm(pathologic_response ~ MEblack + MEblue + MEbrown + MEgreen + MEgreenyellow + MEmagenta + MEpink + MEpurple + MEred + MEsalmon + MEtan + MEturquoise + MEyellow, data=model_data, family=binomial)
summary(m1)

# full model - clinical factors
m2 <- glm(pathologic_response ~ ER_status + tumor_stage + grade + age + PR_status + nodal_status, data=model_data, family=binomial)
summary(m2)

# combined model 
m3 <- glm(pathologic_response ~ MEyellow + ER_status, data=model_data, family=binomial(logit))
summary(m3)

# just yellow eigengene (mostly for plotting)
m4 <- glm(pathologic_response ~ MEyellow, data=model_data, family=binomial(logit))
summary(m4)

# just ER status 
m5 <- glm(pathologic_response ~ ER_status, data=model_data, family=binomial)

```

# Logistic Regression Model Plots
```{r}
# plots 
p <- model_data |> ggplot(aes( 
  x = MEyellow, 
  y = pathologic_response, 
  colour = pathologic_response 
)) + 
  geom_jitter( 
    size = 1, alpha = 0.4, 
    height = 0.05, width = 0.1 
  ) + 
  geom_function( 
    #    data = plot_data, 
    aes(x = MEyellow), 
    inherit.aes = FALSE, 
    fun = \(x) predict(m4, 
                       type = "response", 
                       newdata = tibble(MEyellow = x) 
    ), 
    linewidth = 2, colour = "#0072B2" 
  ) + 
  labs( 
    x = "Team Score", 
    y = "Status y = 1 if Win, 0 otherwise\n and modeled probabilities" 
  ) + 
  scale_y_continuous(breaks = c(0, 0.5, 1)) 
p + theme(legend.position = "none") 


```


# Interpretations
```{r}
# interpret 
predict(m5, type = "response")
table1 <- broom::tidy(m3, exponentiate=TRUE, conf.int=TRUE)
table1
```

# Diagnostics 
```{r}
# model 3 
check_model(m3)

# collinearity
check_collinearity(m3)

# model fit 
result <- binned_residuals(m4)
as.data.frame(result)
plot(result, show_dots=TRUE)

r2(m3)

# cooks distance 

# linearity 
```


# Model Performance
```{r}
# confusion matrix 

# AUC