---
title: "02_model"
author: "Riley Butler"
date: "2026-08-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This file is used for model construction. 

```{r}
# set-up 
if (!dir.exists("results")) {
  dir.create("results", recursive = TRUE)
}

wgcna_path <- "data/wgcna_results.rds"
if (!file.exists(wgcna_path)) {
  system("Rscript 01_wgcna.R")
}

#wgcna_path2 <- "data/wgcna_results-spt4.rds"
#if (!file.exists(wgcna_path2)) {
#  system("Rscript 01_wgcna.R")
#}
```

```{r}
# load libraries 
library(tidyverse)
library(performance)
```

```{r}
# load data
data <- readRDS(wgcna_path)
#data <- readRDS(wgcna_path2)
clinical_factors <- data$pheno
clinical_factors <- clinical_factors |> select(-geo_accession)
eigengenes <- data$eigengenes

# create dataframe
model_data <- data.frame(clinical_factors,eigengenes)

# swapping variables - maybe do this early on? 
model_data$pathologic_response <- ifelse(model_data$pathologic_response == 0, 1, 0)

```

Now that the data is set-up - lets explore! 

# Logistic Regression Models 

These models have assumptions, including 
1. The outcome is binary. In this case, pCR = 1 and RD = 0
2. Linearity of log-odds. This means continuous predictors (MEs, age, etc) are linearly related to the log-odds of the outcome. 
3. No multicolinearity, the predictors aren't strongly correlated. This is measured by VIF, with a VIF less than 10 indicating no colinearity. 

```{r}

# null model 
m0 <- glm(pathologic_response ~ 1, data = model_data, family = binomial)
summary(m0)

# full model - eigengenes
m1 <- glm(pathologic_response ~ MEblack + MEblue + MEbrown + MEgreenyellow + MEmagenta + MEpink + MEred + MEsalmon + MEturquoise + MEyellow, data=model_data, family=binomial)
summary(m1)

# full model - clinical factors
m2 <- glm(pathologic_response ~ ER_status + tumor_stage + grade + age + PR_status + nodal_status, data=model_data, family=binomial)
summary(m2)

# combined model 
m3 <- glm(pathologic_response ~ MEyellow + ER_status + MEpink, data=model_data, family=binomial(logit))
summary(m3)

# just yellow eigengene (mostly for plotting)
m4 <- glm(pathologic_response ~ MEyellow, data=model_data, family=binomial(logit))
summary(m4)

# just ER status 
m5 <- glm(pathologic_response ~ ER_status, data=model_data, family=binomial)
summary(m5)
```

# Logistic Regression Model Plots
```{r}
# plots 
p <- model_data |> ggplot(aes( 
  x = MEyellow, 
  y = pathologic_response, 
  colour = pathologic_response 
)) + 
  geom_jitter( 
    size = 1, alpha = 0.4, 
    height = 0.05, width = 0.1 
  ) + 
  geom_function( 
    #    data = plot_data, 
    aes(x = MEyellow), 
    inherit.aes = FALSE, 
    fun = \(x) predict(m4, 
                       type = "response", 
                       newdata = tibble(MEyellow = x) 
    ), 
    linewidth = 2, colour = "#0072B2" 
  ) + 
  labs( 
    x = "Team Score", 
    y = "Status y = 1 if Win, 0 otherwise\n and modeled probabilities" 
  ) + 
  scale_y_continuous(breaks = c(0, 0.5, 1)) 
p + theme(legend.position = "none") 


```


# Interpretations
```{r}
# interpret 
predictions <- predict(m4, type = "response") # highest probability of pCR is 49% 
table1 <- broom::tidy(m3, exponentiate=TRUE, conf.int=TRUE) # shows the significance of the coefficients 
table1 
```

# Diagnostics 
```{r}
# model 3 
check_model(m4)

# collinearity
check_collinearity(m4)

# model fit 
result <- binned_residuals(m4)
as.data.frame(result)
plot(result, show_dots=TRUE)

r2(m3)

# cooks distance 

# linearity 
```


# Model Performance
```{r}
# confusion matrix 

# AUC

```


############################ Elastic Net Regression ############################ 
```{r}
library(glmnet)
library(caret)
```

```{r}

x <- model_data |> select(-pathologic_response)
x <- as.matrix(x)
y <- model_data$pathologic_response
fit <- glmnet(x, y, family="binomial", label=TRUE, alpha = 0.5125)

plot(fit)

print(fit)
# shows the number of nonzero coefficients, the percent deviance explained and the value of lambda

coef(fit, s=0.01) # shows co-efficients used as lambda = 0.01 

# to choose a model - returns a cv.glmnet object
cvfit <- cv.glmnet(x, y, family="binomial", type.measure="auc", alpha = 0.5125)

plot(cvfit) # plots cross validation curve and upper and lower standard deviation curves along the lambda sequence (error bars). The dotted lines shows special lambda values. 

cvfit$lambda.min # gives minimum mean cross-validated error
cvfit$lambda.1se # gives the value of lambda with the most regularized model (such that the cross-validated error is withing 1 standard error of the minimum)

predict(cvfit, newx = x[1:10, ], s = "lambda.min", type = "class")

```

# Training Model (finding optimum alpha and lambda values)
Elastic net penality controlled by alpha. 
Alpha = 1 is lasso (and default). Picks one of the predictors and discards the others. 
Alpha = 0 is ridge regression. Shrinks the coefficients of correlated predictors towards each other. 

Lambda = tuning paramater that controls the overall strength of the penalty. 



```{r}

X <- model_data |> select(-pathologic_response)
Y <- model_data |> select(pathologic_response)

control <- trainControl(method = "repeatedcv", 
                        number = 5,
                        repeats = 5,
                        verboseIter = TRUE, 
                        classProbs = TRUE,
                        summaryFunction = twoClassSummary)

model_data$pathologic_response <- factor(
  model_data$pathologic_response,
  levels = c(1, 0),
  labels = c("pCR", "RD")
)

elastic_model <- train(pathologic_response ~ ., 
                       data = model_data,
                       method = "glmnet",
                       preProcess = c("center", "scale"), 
                       tuneLength = 25, 
                       trControl = control, 
                       metric = "ROC"
                      
                       )

elastic_model$bestTune # best alpha: 0.5125 
                       # best lambda: 0.0934 

plot(elastic_model)

pred_prob <- predict(elastic_model, X, type = "prob")
pred_prob

elastic_model$results

pred_class <- predict(
  elastic_model,
  newdata = model_data
)

model_data$pathologic_response <- factor(
  model_data$pathologic_response,
  levels = c(1, 0),
  labels = c("pCR", "RD")
)

confusionMatrix(
  pred_class,
  model_data$pathologic_response,
  positive = "pCR"
)
```

```{r}

# looking at correlations between predictors - not super important if using elastic net
cor_matrix <- cor(
  x,
  use = "pairwise.complete.obs",
  method = "pearson"
)

round(cor_matrix, 2)

library(corrplot)

corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  tl.cex = 0.7
)

```


