---
title: "Model Performance"
author: "Riley Butler"
date: "2026-08-30"
output: html_document
---

This file is used to test the model performance by running diagnostic tests, performance checks and making predictions using the test set. 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::opts_knit$set(
  root.dir = normalizePath("..")
)
```

```{r}
# script set-up
model_path <- "data/models.rds"
test_path <- "data/validation_processed_data.rds"
test_MEs_path <- "data/testing_MEs.rds"

if (!file.exists(model_path)) {
  system("Rscript src/02_model_selection.R")
}

if (!file.exists(test_path)) {
  system("Rscript src/00_preprocessing.R TRUE")
}

if (!file.exists(test_MEs_path)) {
  system("Rscript src/03_ME_projection.R")
}

```

```{r}
# load libraries 
library(tidyverse)
library(performance)
library(caret)
library(pROC)
```

```{r}
# load data
data <- readRDS(model_path)
all_test <- readRDS(test_path)
test_cf <- all_test$phenotype
test_MEs <- readRDS(test_MEs_path)

stopifnot(identical(test_cf$geo_accession, rownames(test_MEs)))

full_test_data <- data.frame(test_cf,test_MEs)

# convert to factor for calculating ROC
full_test_data$pathologic_response <- factor(full_test_data$pathologic_response, levels = c(1, 0), labels = c("pCR", "RD"))
```


```{r}
# load data into model types 
model_data <- data$model_data

elastic_model <- data$elastic_model 

ME_logistic_model <- data$ME_logistic_model

CF_logistic_model <- data$CF_logistic_model

full_logistic_model <- data$full_logistic_model

SVM_model <- data$SVM_model

RF_model <- data$random_forest_model

```

## Testing Predictive Power 

#### Do Predictors Improve Predictive Power Compared to the Null? 

```{r}
# testing if predictors contribute to predictive power
null_model <- glm(pathologic_response ~ 1, data = model_data, family = binomial)
logistic1 <- glm(pathologic_response~., data = model_data, family = binomial)
anova(null_model, logistic1, test = "Chisq")

```

The anova test shows that a significant p-value, thus we can reject the null and at least one group mean is statistically different from the others. This means that the predictors improve model fit. 

## Model Comparison 

#### What Model Predicts Chemotherapy Reponse Best? 
```{r}
# comparing models using AUC, Specificity and Sensitivity 
results <- resamples(list(glmnet = elastic_model, MElogistic = ME_logistic_model, CFlogistic = CF_logistic_model, fullLogistic = full_logistic_model, SVM_model=SVM_model, randomForest = RF_model))
summary(results)
box_whisker <- bwplot(results)

box_whisker

saveRDS(box_whisker, "results/box_wisker.rds")
```

The box and whisker plot shows that most metrics are not statistically significantly different across models. The median ROC across models ranges from a minimum of 0.68 for the SVM model and a maximum of 0.78 for the elastic net model. The sensitivity ranges from 0 for the SVM and elastic net model to 0.18 for the ME and full logistic models. The low sensitivity is in part due to the low number of pCR cases in the training set. The specificity ranges from 1 for the elastic net logistic regression model to 0.92 for the full logistic regression model. I chose to look at the elastic net logistic regression model as its construction best fits with the context of the data (i.e. chooses from lots of predictors with alpha and lambda values). The threshold for classification is adjusted from 0.5 using Youden's J Statistic to find a new threshold in order to improve the models sensitivity. 

## Contribution of Module Eigengenes 

#### Do MEs Improve Model Performance? 
```{r}
ME_comparison <- resamples(list(Clinical = CF_logistic_model, ME = ME_logistic_model, Full = full_logistic_model))

summary(ME_comparison)

bwplot(ME_comparison, metric = "ROC")

plot(varImp(full_logistic_model))
```

The results show that including module eigengenes don't significantly improve ROC, sensitivity or specificity (for the logistic regression models). However, predictor importance shows that some module eigengenes are more important than some clinical factors, indicating the importance of including genetic data through networks analysis for predicting chemotherapy response. These results could be compared to using differential abundance of individual genes to show whether conducting a networks analysis significantly improves models using genetic data.  

## Elastic Net Logistic Regression Diagnostics 

```{r}
# shows model performance (ROC-AUC) for different alpha and lambda values
ggplot(elastic_model, xvar="lambda")
```


This plot shows the cross-validated ROC-AUC values across different combinations of alpha and lambda. For elastic net logistic regression, the lambda value controls strength of regularisation whist alpha balances ridge and lasso penalties. The graph shows that performance is not highly sensitive to exact tuning parameters. However, lower alpha values performed best which indicates ridge-dominated penalty was appropriate for the correlated predictors. 

```{r}
elastic_model$bestTune
```
The best tunning paramets are alpha = 0 and lambda = 1

```{r}
densityplot(elastic_model, pch = "*")
```

This plot shows the ROC values across different samples. The most abundant ROC appears to be around ~0.8. The curve shows that there is some variability in performance across samples 

```{r}
# shows the coefficients used for the final model. Note that some correlated coefficients in the model are shrunk. 
coef(elastic_model$finalModel, s = elastic_model$bestTune$lambda)
```
This summary shows the coefficients for the different predictors. None of the predictors appear to have significantly higher coefficient values. 

Because none of the probabilities exceed 0.5, all samples are classified as RD. To improve sensitivity, the threshold is changed using Youden's J Statistic. 

## Adjusting Threshold 

```{r}
# find the average predictions and probabilities for eachs observation 
patient_predictions <- elastic_model$pred |> group_by(rowIndex) |> summarise(obs = first(obs), pCR_prob = mean(pCR), .groups = "drop")

# takes actual outcome for each patient and the models predicted probability, calculates how sensitivity and specificity changes across different cutoffs
roc_obj <- roc(response = patient_predictions$obs, predictor = patient_predictions$pCR_prob, levels = c("RD", "pCR"), direction = "<")
```

```{r}
# plotting the ROC curve
plot(roc_obj)
```

The curve bends to the left, showing some improvement in performance (i.e. sensitivity) against the 'null' (the straight diagonal line)

Finding the best threshold by balancing sensitivity and specificity using Youden-selected threshold, which maximizes: sensitivity + specificity - 1

```{r}
best_threshold <- coords(roc_obj, x = "best", best.method = "youden", ret = c("threshold", "sensitivity", "specificity"))

best_threshold

# assign optimal cutoff value 
cutoff <- as.numeric(best_threshold["threshold"])

patient_predictions$pred <- factor(ifelse(patient_predictions$pCR_prob >= cutoff, "pCR", "RD"))
patient_predictions$obs <- factor(patient_predictions$obs,levels = c("pCR", "RD"))

roc_obj$auc # 0.7668

# make a confusion matrix with the new cutoff 
confusionMatrix(patient_predictions$pred, patient_predictions$obs, positive = "pCR")
```

These results show a sensitivity of 0.77 and specificity of 0.72. This new threshold better balances sensitivity and specificity.
Kappa: 0.313 shows fair-to-moderate agreement beyond chance between predicted classes and the observed pCR/RD classes
Mcnemar's Test P-value: very small, showing that the model is asymmetric (the model makes one type of mistake more than the other)
AUC: 0.7686 

```{r}
# constructing regularization plot 
regularization_plot <- plot(
  elastic_model$finalModel,
  xvar = "lambda",
  label = TRUE
)

# mapping predictors to numbers to interpret the plot
x <- model.matrix(pathologic_response ~ .,data = model_data)[, -1]
colnames(x)
```
The plot shows that predictor coefficients change as elastic-net penalty increases. The plot shows that the important predictors (i.e. predictors with most deviation from 0) are MEblack, tumor stage, MEpurple and ggi_class. As lambda decreases (-log(lambda) increases), the absolute value of predictor coefficients increases. 

```{r}
calibration_data <- patient_predictions |> mutate(actual = ifelse(obs == "pCR", 1, 0), bin = ntile(pCR_prob, 5)) |> group_by(bin) |> summarise(predicted = mean(pCR_prob), observed = mean(actual), n = n(), .groups = "drop")
```

Calibration Plot: shows predicted probability vs observed probability. A perfectly calibrated model falls on the diagonal line (X = Y). This plot shows whether the model overestimates or underestimates pCR across 5 bins (where bins contain samples with similar predicted probabilities). 

```{r}
plot(
  calibration_data$predicted,
  calibration_data$observed,
  xlim = c(0, 1),
  ylim = c(0, 1),
  xlab = "Mean predicted probability of pCR",
  ylab = "Observed proportion with pCR",
  pch = 19
)

abline(0, 1, lty = 2)

lines(
  calibration_data$predicted,
  calibration_data$observed
)
```

The plot shows that the model is well calibrated for predicted probability of 0.2 but over estimates pCR probability for patients with lower probabilities and underestimates for patients with lower pCR probability. This is likely good for distinguishing between pCR and RD in the case of classification. However, the instability makes the model less reliable and less predictable. 

## Making Predictions on Test Set 

```{r}
# ensuring predictors align between validation set and model 
required_predictors <- setdiff(
  names(elastic_model$trainingData),
  ".outcome"
)

missing_predictors <- setdiff(
  required_predictors,
  names(full_test_data)
)

stopifnot(length(missing_predictors) == 0)

test_for_model <- full_test_data |>
  select(all_of(required_predictors))

predictions_testing <- predict(
  elastic_model,
  newdata = test_for_model,
  type = "prob"
)
```



```{r}
predictions_training <- predict(elastic_model, newdata=model_data, type="prob")
head(predictions_training)
```
Most samples have 0.10 to 0.30 probability of being pCR and 0.70 to 0.90 probability of being RD. This generally aligns with the proportions of the training set.

```{r}
# ROC curve with validation set 

test_results <- data.frame(obs = full_test_data$pathologic_response, pCR_prob = predictions_testing$pCR)
test_results$obs <- factor(test_results$obs, levels = c("RD", "pCR"))
test_roc <- roc(response = test_results$obs, predictor = test_results$pCR_prob, levels = c("RD", "pCR"), direction = "<")

test_roc
auc(test_roc)
ci.auc(test_roc)

plot(test_roc, main = "ROC Curve with Test Set")
```

The ROC curve is over the null curve, thus chemotherapy predictions for the elastic net logistic regression model are better than the null model.

```{r}
# confusion matrix for test set
test_results$pred <- factor(
  ifelse(test_results$pCR_prob >= cutoff, "pCR", "RD"),
  levels = c("pCR", "RD")
)

test_results$obs <- factor(
  test_results$obs,
  levels = c("pCR", "RD")
)

test_cm <- confusionMatrix(
  test_results$pred,
  test_results$obs,
  positive = "pCR"
)

test_cm
```
The accuracy of the test set is 68%, the sensitivity on the test set is 52% and the specificity is 73% on the test set. 





